---
title: Basic traffic data from `stats19`
author: Denis Schluppeck
date: '2021-10-13'
slug: basic-traffic-data-from-stats19
categories:
  - traffic
tags:
  - 20mph
  - nottingham
---

To raise awareness about the actual number of traffic accidents in our neighbourhood I have decided to do some number crunching and plotting (since this is what I do a lot of in my day job as a scientist).

I will try to bring out the numbers as clearly as I can (and I will let them speak for themselves). There is no claim for completeness, but if you want to inspect how I went about it the code (in `R`) is published on my `github` page: https://github.com/schluppeck/make-web-20mph-code/


## To start: a map

These are data from the `stats19` database from the 5 year period 2015-2019.

- I have merged the "casualties" and "accidents" tables available from `stats19` to combine information of **who got injured** and **where the accident occurred*

- only cases handled by the "Nottinghamshire" police are included

- this map only includes cases where the speed limit was <= 40mph

- pedestrian casualties are shown in **red**, **cyclists** in magenta, others (such as people in cars) in blue. Pedestrians and cyclists are shown with slighly larger symbols.


```{r include = FALSE}
library(tidyverse)
library(stats19)
library(sf)
theYear = c(2015,2016,2017,2018,2019)
casualties = get_stats19(year = theYear, type = "casualties")
crashes = get_stats19(year = theYear, type = "accidents")
crashes_sf = format_sf(crashes)
vehicles = get_stats19(year = theYear, type = "vehicles", ask = FALSE)

ng = filter(police_boundaries, pfa16nm == "Nottinghamshire")

crashes_ng = crashes_sf[ng, ]
nrow(crashes_sf)
nrow(crashes_ng)

sel = casualties$accident_index %in% crashes_ng$accident_index
casualties_ng = casualties[sel, ]
cas_types = casualties_ng %>% 
  select(accident_index, casualty_type) %>% 
  mutate(n = 1) %>% 
  group_by(accident_index, casualty_type) %>% 
  summarise(n = sum(n)) %>% 
  tidyr::spread(casualty_type, n, fill = 0) 
cas_types$Total = rowSums(cas_types[-1])
cj = left_join(crashes_ng, cas_types, by = "accident_index")


base::setdiff(names(cj), names(crashes_ng))

theme_set(theme_minimal())
```

```{r, echo=FALSE}
library(leaflet)

# roundabout in WB
y= 52.9311215
x=-1.1270815

# only show events where speed limit was <= 40 mph
cj_local <- cj %>% 
  filter(speed_limit <=40)


m <- leaflet(data = cj_local) %>%
  addProviderTiles(providers$Stamen.Toner) %>%  
  # not default OpenStreetMap map tiles
  addCircleMarkers(
      lng = cj_local$longitude, lat = cj_local$latitude,
      #adjust size
      radius = ~if_else(cj_local$Pedestrian==1 |
                          cj_local$Cyclist==1, 8, 4),
      # colour them in
      color = ~case_when(cj_local$Pedestrian==1 ~ "red",
                         cj_local$Cyclist==1 ~ "magenta",
                       TRUE ~ "blue"), # ~pal(speed_limit),
      stroke = TRUE, fillOpacity = 0.8) %>%
   setView(lng = x, lat = y, zoom=13)

m  # show map


```

