---
title: "When do most pedestrians/cyclists get hit?"
author: "Denis Schluppeck"
date: '2021-10-28'
slug: when-do-accidents-happen-updated-data-from-dft
categories: traffic
tags: 20mph
---

Since I first put up [a map of traffic accidents / casualties](https://schluppeck.github.io/make-wb-20mph/2021/10/13/basic-traffic-data-from-stats19/) at the start of October '21, the Department for Transport has updated its openly accessible data. This means that now data for the years between 2016 and 2020 is easily accessible.

I think  knowing **when** accidents are most common is as important as known **where** they occur across our county. In this blog post, I will look at that data.



```{r, include = FALSE}
library(tidyverse)
library(lubridate)
library(stats19)
library(sf)
theYear = c(2016,2017,2018,2019,2020)

# a function to load a stats19 file, but fix lat/lon issues 
# in some files
get_crashes <- function(aYear) {
  d <- get_stats19(aYear, type = "accident") %>% 
    mutate(longitude = as.numeric(longitude),
           latitude = as.numeric(latitude))
}
# same for reading in files. lat/lon gets wrongly parsed
# as char instead of double, so do casting explicitly
read_crashes <- function(aYear, ...) {
  d <- read_accidents(aYear, ...) %>% 
    mutate(longitude = as.numeric(longitude),
           latitude = as.numeric(latitude))
}


# deal with folder locations during render / test
if ("data" %in% list.files()) {
  data_dir = "./data"
} else {
  data_dir = "../../../data"
}

# if data have been cached:
cache_file_name <- "2021-10-29-when-data.RData" 
if (cache_file_name %in% list.files(path=data_dir)) {
  paste("yay - cached data present")
  paste("... loading")
  load(paste0(data_dir, '/', cache_file_name))
} else {
  paste("nay - loading from CSV files")
  
  # download the data - if CSV files already exist in data_dir, 
  # this will not re-download, but still takes a bit of time
  walk(theYear, dl_stats19, 
       type = "casualty", data_dir = data_dir)
  walk(theYear, dl_stats19, 
       type = "accident", data_dir = data_dir)
  walk(theYear, dl_stats19, 
       type = "vehicle", data_dir = data_dir)

  # read in casualties and merge rowwise
  casualties = map_dfr(theYear, read_casualties, 
                     data_dir = data_dir)
  # crashes
  crashes = map_dfr(theYear, read_crashes, 
                       data_dir = data_dir)
  crashes_sf = format_sf(crashes)
  
  # vehicle data not needed, so save time
  # vehicles = map_dfr(theYear, get_stats19, type = "vehicle")
  
  
  ng = filter(police_boundaries, pfa16nm == "Nottinghamshire")
  
  # subset crashes data to NG only
  crashes_ng = crashes_sf[ng, ]

  # subset casualties to NG area
  sel = casualties$accident_index %in% crashes_ng$accident_index
  casualties_ng = casualties[sel, ]
  
  # summarise casualties
  cas_types = casualties_ng %>% 
    select(accident_index, casualty_type) %>% 
    mutate(n = 1) %>% 
    group_by(accident_index, casualty_type) %>% 
    summarise(n = sum(n)) %>% 
    tidyr::spread(casualty_type, n, fill = 0) 
  cas_types$Total = rowSums(cas_types[-1])
  
  # join casualties and crash data / time, location etc,
  cj = left_join(crashes_ng, cas_types, by = "accident_index")
}


# set theme for plotting
theme_set(theme_minimal())
```

## When do most accidents happen during the day / week?

A look at the pattern of accidents in Nottinghamshire (not just West Bridgford) **over time** is quite revealing - it provides quite a bit of additional insight to what we already learnt from [the map view](https://schluppeck.github.io/make-wb-20mph/2021/10/13/basic-traffic-data-from-stats19/) from a previous post. 

In particular, let's look at the average pattern of pedestrian and cyclist casualties during the day and across the week (over the last 5 years).



```{r, echo = FALSE, message = FALSE, warning=FALSE}

# get hour, weekday, week, month and year from datetime object
# keep accident_severity for plotting later.
crash_times = cj %>%
  filter(Pedestrian > 0 | Cyclist > 0) %>% 
  st_set_geometry(NULL) %>% 
  mutate(hour = hour(datetime),
         wday = wday(datetime),
         week = week(datetime),
         month = month(datetime),
         year = year(datetime),
         accident_severity = as.factor(accident_severity)) 

# for heatmap, summarise numbers, still partitioned by severity
ct_grid <- crash_times %>% 
  group_by(hour, wday) %>% 
  summarise(fatal = sum(accident_severity == "Fatal"),
             serious = sum(accident_severity == "Serious"),
            slight = sum(accident_severity == "Slight")) 

# another version that's gridded by month year 
# not currently used for plotting, though
ct_grid_week <- crash_times %>% 
  group_by(month, year) %>% 
  summarise(fatal = sum(accident_severity == "Fatal"),
             serious = sum(accident_severity == "Serious"),
            slight = sum(accident_severity == "Slight")) 

```

```{r, echo = FALSE, message=FALSE, warning=FALSE}

## this is the day / week plot
#Assign color variables
col1 = "#d8e1cf"  # gray 
col2 = "#ff2222"  # reddish

g <- ct_grid %>% ggplot( aes(hour, wday)) + 
  geom_tile(aes(fill = slight+fatal+serious),
            colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) + 
  scale_x_continuous(breaks = c(6,9,12,15,18)) +
  scale_y_reverse(breaks = 1:7, labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")) +
  guides(fill=guide_legend(title="# casualties")) +
  labs(title = "Nottinghamshire accidents",
       subtitle = "2016-2020, data: dft.gov.uk",
       x = "Time of day" ,
       y = "",) +
  geom_vline(xintercept = c(9,15), color = "black", size=0.75) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), aspect.ratio = 0.6, text = element_text(size=18)) 

g
```

- the heatmap plot shows more casualties on weekdays around 9am and just after 3pm - the times when lots of people are on the road -- commuting to work and getting to and from school: the more red the colour, the more accidents happened. Each row corresponds to a weekday (top: Sunday) and each column to a one-hour slot during the day (starting at 0000h on the left).

- maybe not surprisingly, weekdays are worse than Saturday/Sunday


## What age groups are affected when?

One important question for me was: 

>Are there times of day that are more dangerous than others? What do the data from the last 5 years show?

The distribution of pedestrian and cyclist casualties over the the span of the day, also **taking into account the age group**, shows a very clear pattern:


```{r, echo = FALSE, warning=FALSE,message=FALSE, out.height="50%"}

# summarise casualties / only pedestrians, cyclists...
# the age bands in DFT data are a bit narrow, resulting in 
# maybe too many plots to be useful
# I first turn age_bands into an ordered factor
# then re-bin the data to <20, 21-55, >56, and missing

age_bands = c("0 - 5", "6 - 10","11 - 15","16 - 20","21 - 25", "26 - 35", "46 - 55", "56 - 65", "66 - 75",  "Over 75",  "Data missing or out of range")

cas_types_age = casualties_ng %>% 
  select(accident_index, casualty_type, age_band_of_casualty) %>%
  mutate(age_band_of_casualty = ordered(age_band_of_casualty,
                                        levels=age_bands),
         age_group = fct_collapse(age_band_of_casualty,
            `<20y` = c("0 - 5", "6 - 10","11 - 15","16 - 20"),
            `21-55y` = c("21 - 25", "26 - 35", "46 - 55"),
            `>56y` = c("56 - 65", "66 - 75",  "Over 75"),
            missing = c("Data missing or out of range")),
         age_group = fct_explicit_na(age_group, "missing"),
         n = 1) %>% 
    group_by(accident_index, age_group, .drop=FALSE) %>% 
    group_by(casualty_type, .add=TRUE) %>% 
    summarise(n = sum(n)) 
  
# join casualties and crash data / time, location etc,
cj_age = left_join(crashes_ng, cas_types_age, by = "accident_index")

age_plot <- cj_age %>% 
  st_set_geometry(NULL) %>% 
  filter(casualty_type %in% c("Pedestrian", "Cyclist")) %>% 
  mutate(wday = wday(datetime),
         month = month(datetime),
         hour = hour(datetime)) %>% 
  ggplot(aes(x = hour, fill = casualty_type )) + 
  geom_histogram(bins = 24) +
  scale_fill_brewer(palette = "Set1",guide="none") +
  facet_grid(age_group~casualty_type) +
  scale_x_continuous(breaks = c(6,9,12,15,18)) +
  scale_y_continuous(breaks = c(0, 50, 100)) +
  labs(title = "Nottinghamshire accidents (by daytime, age)",
       subtitle = "2016-2020, data: dft.gov.uk",
       x = "Time of day" ,
       y = "Count") +
  geom_vline(xintercept = c(9,15), color = "black", size=0.5) +
  theme( panel.grid.minor = element_blank(),
         text = element_text(size=18))

age_plot
```

- the largest number of accidents occur **around 9am** (900h) and **around 3pm** (1500h): there are peaks in the histogram plots close to those times - I have added vertical lines to label 9am and 3pm.

- especially in the afternoon (3pm onwards) a **large number of those younger than 20y** are affected: the peaks in the first row of plots (cyclists on the left, pedestrians on the right) are much larger.

- the peak times for accidents in 21 - 55 year olds is shifted to slighly later. This presumably reflects the later commute times compared to school finishing times.

**Food for thought!**

## Reality check: 2020 & COVID?

For the map shown below, data from the `stats19` database now represent the 5 year period of 2016-2020.

- as before, I have merged the "casualties" and "accidents" tables available from `stats19` to combine information of **who got injured** and **where the accident occurred**.

- key: pedestrian casualties in **red**, cyclists in **magenta**, others (such as people in cars) in **blue**.


```{r, echo=FALSE}
library(leaflet)

# roundabout in WB
y= 52.9311215
x=-1.1270815

# only show events where speed limit was <= 40 mph
cj_local <- cj %>% 
  filter(speed_limit <=40)

m <- leaflet(data = cj_local) %>%
  addProviderTiles(providers$Stamen.Toner) %>%  
  # not default OpenStreetMap map tiles
  addCircleMarkers(
      lng = cj_local$longitude, lat = cj_local$latitude,
      #adjust size
      radius = ~if_else(cj_local$Pedestrian==1 |
                          cj_local$Cyclist==1, 4, 2),
      # colour them in
      color = ~case_when(cj_local$Pedestrian==1 ~ "red",
                         cj_local$Cyclist==1 ~ "magenta",
                       TRUE ~ "blue"), # ~pal(speed_limit),
      stroke = TRUE, fillOpacity = 0.8) %>%
   setView(lng = x, lat = y, zoom=13)

m  # show map
```

