---
title: "Results of informal survey"
author: "R package build"
date: '2021-10-15'
output:
  html_document:
    df_print: paged
categories: traffic
tags:
- 20mph
- wb
slug: results-of-informal-survey
---

Together with my original facebook post, I put together a [very short survey](https://forms.gle/TigMEgcqZXq5vApm7) and asked the following questions:

1. "What do you think about making the speed limit on all residential roads in WB 20mph?"

2. "Any other ideas / comments / questions?"

3. "Your e-mail address (optional)" - for letting people know what I found out.

An important caveat here is that this is not a scientific poll and just because there are lots of positive views expressed here doesn't necessarily mean that the same proportion of people in all of WB feel this way.

Nevertheless, I think it's clear that there is a **strong** desire for some change.

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(kableExtra)

# enable rendering from within RStudio console
# where getwd() is different that in rendering context
if ("data" %in% list.files()) {
  pathvar = "./"
} else {
  pathvar = "../../../"
}

d <- readxl::read_excel(str_c(pathvar, "data/make-wb-20mph-spreadsheet.xlsx")) %>% 
  clean_names() %>% 
  rename(in_favour = starts_with("what_do_you"),
         other_thoughts = starts_with("any_other")) %>% 
  select(-1, -last_col())
  
theme_set(theme_minimal())
```

## Summary

As of today (2021-10-15), 36 people filled in the form.

```{r, echo=FALSE}
d %>% 
  ggplot(aes(x = for_against, fill = for_against)) + 
  geom_bar(show.legend = FALSE) +
  scale_fill_brewer() +
  geom_text(aes(label = ..count..), stat="count", nudge_y = -2 ) +
  theme(aspect.ratio = 0.5) +
  labs(y = NULL, x = NULL, title = "20mph in WB?") +
  coord_flip() 
    
```


Reading through the comments a couple of things stand out from the responses:

- **enforcement** is felt to be a huge issue.
- there are several place across WB that cause people issues
- **speeding is not the only issue**. it sounds like cars run pedestrian crossings all the time. not cool.
- **speed bumps** may be a bad idea (bad for buses, cars, may encourage speeding between bumps...)

Note: there are some cogent arguments against introducing blanket 20mph. Some of them are expressed in one of the **against** entries. But unfortunately, there is also a favourite "toys-out-of-the-pram" entry in the **against** bracket: 

>Teach your kids to cross the road properly if you’re worried. 

Anonymous of course. Maybe what they meant was for there to be "more education on road safety at home and at school"?. But clearly, some people have deep-seated issues they need to resolve.


## Replies to Question 1 (yes/no?)

```{r, echo=FALSE}
d %>% 
  select(in_favour) %>% 
  rownames_to_column() %>% 
  kbl(escape = F,col.names = NULL) %>% 
  row_spec(which(d$for_against=="for"), background="#EEEEEE") 

```

## Replies to Question 2 (Any other comments)

```{r, echo=FALSE}
d %>% 
  select(other_thoughts) %>% 
  rownames_to_column() %>% 
  kbl(escape = F,col.names = NULL) %>% 
  row_spec(which(d$for_against=="for"), background="#EEEEEE") 

```

